A Linear Solution of the Linearity Challenge

### 2025-12-24 paper/Congruence.lagda.tex
british spelling

We formalize structural precongruence as a binary relation between processes
that are well typed in the \emph{same} typing context. This entails that
structural precongruence preserves typing by definition.

@@

Finally, in \Cref{sec:semantics} we have colloquially defined $\pcong$ as a
``\emph{pre-congruence}'', implying that it is reflexive, transitive relation
and that it is is preserved by some forms of calculus. In the formalization we
have to be precise and we introduce specific rules:

>>>>

We formalise structural precongruence as a binary relation between processes
that are well typed in the \emph{same} typing context. This entails that
structural precongruence preserves typing by definition.

@@

Finally, in \Cref{sec:semantics} we have colloquially defined $\pcong$ as a
``\emph{pre-congruence}'', implying that it is reflexive, transitive relation
and that it is is preserved by some forms of calculus. In the formalisation we
have to be precise and we introduce specific rules:

### 2025-12-24 paper/Context.lagda.tex
british spelling

We are going to adopt a nameless representation of channels hence typing
contexts are represented using lists of types, where the (polymorphic)
\AgdaDatatype{List} type and its constructors \AgdaInductiveConstructor{[]} and
\AgdaInductiveConstructor{\_∷\_} are defined in the module
\AgdaModule{Data.List} of Agda's standard library. We will keep using
$\ContextC$, $\ContextD$ and $\ContextE$ to range over typing contexts as we
have done in \Cref{sec:typing-rules}, even though in the Agda formalization they
are lists instead of finite maps.

@@

Context splitting is going to be pervasive in the subsequent development, to the
point that it contributes substantially to the complexity of the formalization.
For this reason, we invest some effort in defining a few abstractions that allow
us to hide context splittings and enhance code readability.
%
Inspired by separation logic~\cite{OHearnPym99} and
following~\citet{RouvoetPoulsenKrebbersVisser20}, we define the \emph{separating
conjunction} $P \sand Q$ of two predicates $P$ and $Q$ over contexts:
\begin{code}
data _∗_ (P Q : Pred Context _) (Γ : Context) : Set where
  _⟨_⟩_ : ∀{Δ Θ} → P Δ → Γ ≃ Δ + Θ → Q Θ → (P ∗ Q) Γ
\end{code}

The predicate $P \sand Q$ holds for those contexts $\ContextC$ that can be split
into $\ContextD$ and $\ContextE$ where $P$ holds for $\ContextD$ and $Q$ holds
for $\ContextE$.
%
The constructor \AgdaInductiveConstructor{\_⟨\_⟩\_} has three explicit arguments
witnessing the splitting $\Splitting\ContextC\ContextD\ContextE$ along with
$P~\ContextD$ and $Q~\ContextE$.
%
The use of the metavariables $P$ and $Q$ for denoting predicates over contexts
is not without reason. As we will see shortly, in our formalization processes
are indeed predicates over typing contexts, albeit not the only ones.

>>>>

We are going to adopt a nameless representation of channels hence typing
contexts are represented using lists of types, where the (polymorphic)
\AgdaDatatype{List} type and its constructors \AgdaInductiveConstructor{[]} and
\AgdaInductiveConstructor{\_∷\_} are defined in the module
\AgdaModule{Data.List} of Agda's standard library. We will keep using
$\ContextC$, $\ContextD$ and $\ContextE$ to range over typing contexts as we
have done in \Cref{sec:typing-rules}, even though in the Agda formalisation they
are lists instead of finite maps.

@@

Context splitting is going to be pervasive in the subsequent development, to the
point that it contributes substantially to the complexity of the formalisation.
For this reason, we invest some effort in defining a few abstractions that allow
us to hide context splittings and enhance code readability.
%
Inspired by separation logic~\cite{OHearnPym99} and
following~\citet{RouvoetPoulsenKrebbersVisser20}, we define the \emph{separating
conjunction} $P \sand Q$ of two predicates $P$ and $Q$ over contexts:
\begin{code}
data _∗_ (P Q : Pred Context _) (Γ : Context) : Set where
  _⟨_⟩_ : ∀{Δ Θ} → P Δ → Γ ≃ Δ + Θ → Q Θ → (P ∗ Q) Γ
\end{code}

The predicate $P \sand Q$ holds for those contexts $\ContextC$ that can be split
into $\ContextD$ and $\ContextE$ where $P$ holds for $\ContextD$ and $Q$ holds
for $\ContextE$.
%
The constructor \AgdaInductiveConstructor{\_⟨\_⟩\_} has three explicit arguments
witnessing the splitting $\Splitting\ContextC\ContextD\ContextE$ along with
$P~\ContextD$ and $Q~\ContextE$.
%
The use of the metavariables $P$ and $Q$ for denoting predicates over contexts
is not without reason. As we will see shortly, in our formalisation processes
are indeed predicates over typing contexts, albeit not the only ones.

### 2025-12-24 paper/DeadlockFreedom.lagda.tex
british spelling

As we have seen in \Cref{sec:calculus}, the deadlock freedom result
(\Cref{thm:df}) rests on some notions and predicates about processes which we
now formalize in Agda. First of all we need to define the notion of
\emph{thread}, that is any process other than a cut. As it turns out, it is
convenient to provide a more fine-grained classification of threads,
distinguishing between links, input/output actions, and sometimes also whether
such actions operate on free or bound channels. We define predicates for each of
these process classes:

>>>>

As we have seen in \Cref{sec:calculus}, the deadlock freedom result
(\Cref{thm:df}) rests on some notions and predicates about processes which we
now formalise in Agda. First of all we need to define the notion of
\emph{thread}, that is any process other than a cut. As it turns out, it is
convenient to provide a more fine-grained classification of threads,
distinguishing between links, input/output actions, and sometimes also whether
such actions operate on free or bound channels. We define predicates for each of
these process classes:

### 2025-12-24 paper/Examples.lagda.tex
examples in subsection
british spelling

\section{Examples}
\label{sec:examples}

@@

In this section we revisit and expand the sample processes discussed in
\Cref{ex:booleans,ex:echo} and show their encoding in our formalization. The
encoding of $\Bool$ is straightforward.

@@

Now if we ask Agda to normalize the goal
$\AgdaFunction{eval}~(\AgdaFunction{False}~\AgdaFunction{||}~\AgdaFunction{False})$
we obtain
$\AgdaInductiveConstructor{select}~(\AgdaInductiveConstructor{ch~⟨~<~•~⟩~inj₂}~(\AgdaInductiveConstructor{close~ch}))$,
that is the definition of \AgdaFunction{False}, as expected.

@@

From these examples it should be clear that writing \Calculus processes in Agda
is not entirely straightforward and requires a little effort. The lack of names
(for channels) clearly obfuscates the code, but the context splittings that must
be provided are usually small and fairly easy to understand. Most importantly,
the use of \AgdaFunction{≫} helps truncating context splittings as soon as the
seeked channel has been reached.
%
Either way, the formalization is meant to study the meta-theory of \Calculus
rather than implementing a user-friendly DSL for it.

>>>>

\subsection{Examples}
\label{sec:examples}

@@

In this section we revisit and expand the sample processes discussed in
\Cref{ex:booleans,ex:echo} and show their encoding in our formalisation. The
encoding of $\Bool$ is straightforward.

@@

Now if we ask Agda to normalise the goal
$\AgdaFunction{eval}~(\AgdaFunction{False}~\AgdaFunction{||}~\AgdaFunction{False})$
we obtain
$\AgdaInductiveConstructor{select}~(\AgdaInductiveConstructor{ch~⟨~<~•~⟩~inj₂}~(\AgdaInductiveConstructor{close~ch}))$,
that is the definition of \AgdaFunction{False}, as expected.

@@

From these examples it should be clear that writing \Calculus processes in Agda
is not entirely straightforward and requires a little effort. The lack of names
(for channels) clearly obfuscates the code, but the context splittings that must
be provided are usually small and fairly easy to understand. Most importantly,
the use of \AgdaFunction{≫} helps truncating context splittings as soon as the
seeked channel has been reached.
%
Either way, the formalisation is meant to study the meta-theory of \Calculus
rather than implementing a user-friendly DSL for it.

### 2025-12-24 paper/Process.lagda.tex
british spelling

We adopt an \emph{intrinsically-typed} representation of processes with
\emph{nameless} channels. The intrinsically-typed representation makes sure that
only well-typed processes can be constructed. This choice increases the effort
in the definition of the datatypes for representing processes and their
operational semantics, but pays off in the rest of the formalization for at
least three reasons:
\begin{itemize}
\item we need not give explicit names to channels, thus we avoid all the
  technicalities and pitfalls that a named representation entails;
\item we conflate processes and typing rules in the same datatype, thus reducing
  the overall number of datatypes defined in the formalization;
\item the typing preservation results are embedded in the very definition of the
  operational semantics of processes and need not require separate proofs
  (\Cref{sec:congruence-agda,sec:reduction-agda}).
\end{itemize}

Channels are not given any name. Instead, they are represented as terms
witnessing that their type is present in the typing context. This is known as
\emph{co-de Bruijn syntax}~\cite{McBride18}, according to which the typing
context associated with a term (a process, a channel) only contains the types of
the channels that actually occur within the term. For this reason, typing
contexts are split as soon as possible, according to the structure of processes,
to make sure that channels are appropriately (and above all linearly)
distributed among sub-processes, so that each channel occurs (and is used)
exactly once.\footnote{Strictly speaking, the typing context of a
\AgdaInductiveConstructor{fail} process may contain channels that do not occur
(and therefore are not used) by the process, but we consider this case as an
exception that does not alter in a substantial way the provided intuition
underlying the use of co-de Bruijn representation.}
%
Concretely, a channel of type $A$ is a predicate that holds for the singleton
context $\AgdaFunction[~A~\AgdaFunction]$:

>>>>

We adopt an \emph{intrinsically-typed} representation of processes with
\emph{nameless} channels. The intrinsically-typed representation makes sure that
only well-typed processes can be constructed. This choice increases the effort
in the definition of the datatypes for representing processes and their
operational semantics, but pays off in the rest of the formalisation for at
least three reasons:
\begin{itemize}
\item we need not give explicit names to channels, thus we avoid all the
  technicalities and pitfalls that a named representation entails;
\item we conflate processes and typing rules in the same datatype, thus reducing
  the overall number of datatypes defined in the formalisation;
\item the typing preservation results are embedded in the very definition of the
  operational semantics of processes and need not require separate proofs
  (\Cref{sec:congruence-agda,sec:reduction-agda}).
\end{itemize}

Channels are not given any name. Instead, they are represented as terms
witnessing that their type is present in the typing context. This is known as
\emph{co-de Bruijn syntax}~\cite{McBride18}, according to which the typing
context associated with a term (a process, a channel) only contains the types of
the channels that actually occur within the term. For this reason, typing
contexts are split eagerly, according to the structure of processes, to make
sure that channels are appropriately (and above all linearly) distributed among
sub-processes, so that each channel occurs (and is used) exactly
once.\footnote{Strictly speaking, the typing context of a
\AgdaInductiveConstructor{fail} process may contain channels that do not occur
(and therefore are not used) by the process, but we consider this case as an
exception that does not alter in a substantial way the provided intuition
underlying the use of co-de Bruijn representation.}
%
Concretely, a channel of type $A$ is a predicate that holds for the singleton
context $\AgdaFunction[~A~\AgdaFunction]$:

### 2025-12-24 paper/Reduction.lagda.tex
british spelling

Just like structural precongruence, reduction is formalized as a binary relation
between processes that are well typed in the very same typing context. Thus, the
definition of reduction embeds the subject reduction result.

>>>>

Just like structural precongruence, reduction is formalised as a binary relation
between processes that are well typed in the very same typing context. Thus, the
definition of reduction embeds the subject reduction result.

### 2025-12-24 paper/Safety.lagda.tex
british spelling

We have seen that type safety is a simple instance of deadlock freedom, which is
made even simpler to formalize in our development where processes are
intrinsically typed. We start by defining reduction contexts as processes with a
single hole. In our intrinsically-typed formalization, reduction contexts are
parameterized by the typing context $\ContextD$ of the hole, which is invariant,
and indexed by the typing context $\ContextC$ of the whole reduction context:

>>>>

We have seen that type safety is a simple instance of deadlock freedom, which is
made even simpler to formalise in our development where processes are
intrinsically typed. We start by defining reduction contexts as processes with a
single hole. In our intrinsically-typed formalisation, reduction contexts are
parameterised by the typing context $\ContextD$ of the hole, which is invariant,
and indexed by the typing context $\ContextC$ of the whole reduction context:

### 2025-12-24 paper/Type.lagda.tex
british spelling

This property is so important in the rest of the formalization that it is worth
defining an implicit rewriting rule so that Agda can autonomously apply it
whenever possible. This is achieved by means of the following pragma
directive.\footnote{The directive is effective provided that the option
\texttt{{-}{-}rewriting} is enabled, either globally when invoking Agda or in
the module's preamble.}

@@

It is worth looking at one case in the proof of \AgdaFunction{dual-subst},
namely when the type is a dualized type variable:

@@

Just like \AgdaFunction{dual-inv}, \AgdaFunction{dual-subst} too is key in the
formalization that follows. Therefore, we add it to the set of implicit
rewriting rules used by Agda so that we do not have to think about this property
again.

>>>>

This property is so important in the rest of the formalisation that it is worth
defining an implicit rewriting rule so that Agda can autonomously apply it
whenever possible. This is achieved by means of the following pragma
directive.\footnote{The directive is effective provided that the option
\texttt{{-}{-}rewriting} is enabled, either globally when invoking Agda or in
the module's preamble.}

@@

It is worth looking at one case in the proof of \AgdaFunction{dual-subst},
namely when the type is a dualised type variable:

@@

Just like \AgdaFunction{dual-inv}, \AgdaFunction{dual-subst} too is key in the
formalisation that follows. Therefore, we add it to the set of implicit
rewriting rules used by Agda so that we do not have to think about this property
again.

### 2025-12-24 paper/conclusion.tex
british spelling

We have presented a formalization of \LCP, a variant of the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99} supported by the same type system
of \CP~\cite{Wadler14}. Binary sessions can be modeled in \LCP using the
continuation-passing encoding described by \citet[extended
version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}.
%
Under the assumption that the size of a formalization is a good measure of its
complexity, the formalization of \LCP can be qualified as ``simple'' compared to
the existing ones.
%
In this respect, our work casts some doubts on the actual contribution of
context splitting to the complexity of a formalization and hints instead at
explicit continuations in the underlying model as a simplifying factor.
%
It is quite interesting to note that, among the approaches that have been
proposed to overcome context splitting, the one of \citet{SanoKavanaghPientka23}
makes use of explicit continuations.

@@

The simple formalization of \LCP is a good basis for further investigations. We
have already extended \LCP with support for infinite types and recursive
processes (the formalization of this extension is publicly available in a
sibling folder of the one for \LCP). In the future, it would be interesting to
formalize the strong normalization property of \LCP as a consequence of cut
elimination of classical linear logic.

In this work we have focused on models and theories of \emph{binary sessions}
(those connecting exactly two processes), but there are also formalizations of
\emph{multiparty sessions}, notably those of \citet{JacobsBalzerKrebbers22a} and
\citet{TiroreBengstonCarbone25}, which are significantly more complex than those
of binary sessions. Just to give an example, the formalization by
\citet{TiroreBengstonCarbone25} amounts to more than 1Mb of Coq code.
Incidentally, multiparty sessions are not supported by a strong connection with
linear logic and there is no known continuation-passing encoding of multiparty
sessions. Hence, it is unclear whether the considerations made for this work
also apply to the multiparty case.

>>>>

We have presented a formalisation of \LCP, a variant of the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99} supported by the same type system
of \CP~\cite{Wadler14}. Binary sessions can be modeled in \LCP using the
continuation-passing encoding described by \citet[extended
version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}.
%
Under the assumption that the size of a formalisation is a good measure of its
complexity, the formalisation of \LCP can be qualified as ``simple'' compared to
the existing ones.
%
In this respect, our work casts some doubts on the actual contribution of
context splitting to the complexity of a formalisation and hints instead at
explicit continuations in the underlying model as a simplifying factor.
%
It is quite interesting to note that, among the approaches that have been
proposed to overcome context splitting, the one of \citet{SanoKavanaghPientka23}
makes use of explicit continuations.

@@

The simple formalisation of \LCP is a good basis for further investigations. We
have already extended \LCP with support for infinite types and recursive
processes (the formalisation of this extension is publicly available in a
sibling folder of the one for \LCP). In the future, it would be interesting to
formalise the strong normalisation property of \LCP as a consequence of cut
elimination of classical linear logic.

In this work we have focused on models and theories of \emph{binary sessions}
(those connecting exactly two processes), but there are also formalisations of
\emph{multiparty sessions}, notably those of \citet{JacobsBalzerKrebbers22a} and
\citet{TiroreBengstonCarbone25}, which are significantly more complex than those
of binary sessions. Just to give an example, the formalisation by
\citet{TiroreBengstonCarbone25} amounts to more than 1Mb of Coq code.
Incidentally, multiparty sessions are not supported by a strong connection with
linear logic and there is no known continuation-passing encoding of multiparty
sessions. Hence, it is unclear whether the considerations made for this work
also apply to the multiparty case.

### 2025-12-24 paper/related.tex
british spelling

We have compared our formalization with others that have appeared in the
literature and for which we were able to retrieve the source code. We have
looked at those formalizations that include the following features:
\begin{itemize}
\item reduction/evaluation semantics for a calculus with support for binary
    sessions, either natively or through their encoding in terms of linear
    channels~\cite{KobayashiPierceTurner99,Kobayashi02b};
\item a type system along with the usual typing preservation results;
\item optionally, a proof of global progress/deadlock freedom.
\end{itemize}

\citet{GotoEtAl16} describe the formalization of a session-based variant of the
$\pi$-calculus which supports channel polymorphism. This is the oldest
formalization of a session-based calculus for which we were able to retrieve the
source code (the work of \citet{Gay01} predates this one, but its source code is
not publicly available any more).

\citet{Thiemann19} formalizes a subset of \GV~\citep{GayVasconcelos10}, a
functional language extended with session communication primitives, along with
an interpreter. \citet{CicconePadovani20} have taken inspiration from Thiemann's
work to formalize a variant of the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99} that supports dependent types, so
as to enable the description of communication protocols whose structure depends
on the content of messages.

\citet{CastroFerreiraYoshida20} describe EMTST, a library for the formalization
of session type systems which includes as case studies the session calculus of
\citet{HondaVasconcelosKubo98} (called ``original system'') and a revised
version of it that is more amenable to be formalized using a locally nameless
representation of channels.

\citet{RouvoetPoulsenKrebbersVisser20} present a library of abstractions
inspired to separation logic aiding the formalization of interpreters for
computing models supported by linear type systems. One of the presented case
studies is the formalization of a fragment of \GV~\citep{Wadler14}. The
separating conjunction we have introduced in \Cref{sec:context-agda} and the
typing of the constructors for the representation of processes in
\Cref{sec:process-agda} have been heavily inspired by this work.

% as case study for their approach based on separation logic, which we have
% partially taken inspiration from for the representation of processes
% (\cref{sec:context-agda,sec:process-agda}). Their formalization adopts the same
% intrinsically-typed representation of processes with nameless channels as our
% own, but their formalization focuses on the implementation of an interpreter (no
% small-step semantics is defined) and subject reduction is

All the formalizations mentioned so far make use of context splitting. In
contrast, \citet{ZalakainDardha21} formalize a generalization of the linear
$\pi$-calculus which is parametric in a \emph{usage algebra} (to account for
channel sharing/linearity) and that is based on \emph{leftover
typing}~\cite{Allais17}. 
%
With leftover typing a process is typed with respect to an \emph{input context},
which describes the overall available channels, and an \emph{output context},
which describes the residual channels not consumed by the process. In this way,
it is possible to ``concatenate'' typing judgments instead of splitting typing
contexts beforehand. As \citet{ZalakainDardha21} nicely summarize, ``context
splits contain usage information that is already present'' in processes, and
therefore can be omitted. This is true, but only provided that channels are
given a \emph{name} in processes (\citet{ZalakainDardha21} use de Bruijn indices
to this aim). In fact, the co-de Bruijn representation of
processes~\cite{McBride18}, whereby channels are nameless and context splitting
is performed eagerly, can be seen as the ``dual approach'' of leftover typing:
channel names provide information that is already present in their (singleton)
typing context, hence they can be omitted from contexts and processes.

Another work based on context splitting is the one by
\citet{JacobsBalzerKrebbers22}, who formalize a library of \emph{connectivity
graphs} for reasoning and enforcing deadlock freedom in a variant of
\GV~\citep{GayVasconcelos10,Wadler14}. This is the first formalization of
deadlock freedom for a calculus of sessions.

@@

    \caption{\label{tab:comparison} A summary of different formalizations of binary session calculi. All sizes are in kb.}

@@

        \textbf{Formalization} &

@@

\Cref{tab:comparison} summarizes the comparison of our formalization with the
aforementioned ones. The three leftmost colums list the reference papers (sorted
by publication date), the proof assistant and the formalized calculus.
%
For simplicity, we have grouped the calculi into classes identified by an
acronym. Some acronyms, like \CP~\cite{Wadler14} and
\GV~\cite{GayVasconcelos10,Wadler14} are well known in the literature on session
types. It should be noted that the actual calculus being formalized usually
differs from (typically, is a strict subset of) the one identified by the
acronym, but it retains at least some of its key features. We use \SessionPI to
denote (variants of) the session-based $\pi$-calculus presented by
\citet{HondaVasconcelosKubo98} and \LinearPI to denote (variants of) the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99}, into which binary sessions can be
encoded using explicit continuation channels as described by \citet[extended
version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}. Finally, we use
\SCP for the structural version of \CP introduced by
\citet{SanoKavanaghPientka23} and \Calculus for our calculus, a linear version
of \CP with explicit continuations.

The four middle columns of \Cref{tab:comparison} report the relevant qualitative
aspects of the formalizations, namely the technique used to enforce the linear
usage of channels, the technique used to encode channels, whether the
formalization is intrinsically or extrinsically typed, and whether the calculus
makes use of continuation channels. Concerning the use of continuations, the
approaches based directly on the linear $\pi$-calculus (into which sessions can
be encoded) are marked with {\YES} and those based on a calculus with native
sessions are marked with {\NO}. The calculus \SCP is marked with {\MAYBE}
because, while not being strictly based on the linear $\pi$-calculus, it makes
use of explicit continuations for defining the linearity predicates that check
the linear usage of channels. Finally, all the approaches based on \GV are also
marked with \MAYBE. In principle, \emph{none} of these calculi makes use of
continuations. However, \GV is designed in such a way that each session
operation acting on a channel $s$ returns the result of the operation (if any)
\emph{along with the same channel $s$}. In this way, the type of $s$ can be
``updated'' to take into account the effect of the operation. As observed by
\citet{Padovani17}, this way of modeling the communication primitives is
virtually indistinguishable from one making use of explicit continuation
channels.

The rightmost columns of \Cref{tab:comparison} report the \emph{size} (in
kilobytes) of the formalizations as rough (and possibly questionable) estimates
of their complexity. Instead of counting line numbers, which may be affected by
code indentation styles and syntactical constraints of the proof assistant being
used, we have counted the number of characters after comments have been removed
and spaces have been squeezed.\footnote{Sequences of two or more consecutive
space-like characters are collapsed into a single space. The squeezing is
obtained by running the command \texttt{tr -s [:space:] file} on Unix-like
systems.}
%
The reported sizes account for the source code of the formalization excluding
examples and any safety result, if present. We have noticed that the
interpretation of ``safety'' varies widely among the formalizations and, except
for our own, they all differ from the one stated in the linearity
challenge~\citep{CarboneEtAl24} when available.
%
Some formalization define \emph{libraries} which could be reused in different
contexts. In these cases, the size of the library is reported separately next to
the size of the part of the development that uses it.

In general, it is difficult to draw firm conclusions on the effectiveness of the
various approaches in addressing the linearity challenge because the
formalizations differ widely for a variety of entangled factors including the
reference calculus, supported features, proof assistant,
process/expression/channel representation, handling of linearity. Also, most
formalizations predate the publication of the linearity
challenge~\cite{CarboneEtAl24}, hence they are not necessarily targeted to
solving that particular problem. However, some of
them~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,SanoKavanaghPientka23,ZackonEtAl25}
are explicitly motivated by the complexity arising from the linear usage of
resources/channels and from context splitting in particular.

Based on the available data, we can make the following observations.
%
The use of context splitting, which is very well represented, does not seem to
be a good indicator of the complexity of the formalization, since it spans the
whole range of sizes. Our formalization is based on context splitting and it is
the only one that covers \CP in full, yet it is by far the smallest one.

The largest formalizations~\cite{GotoEtAl16,CastroFerreiraYoshida20} are also
the ones that adopt a locally nameless representation of channels. In this
formalizations channels are represented in two different ways, depending on
whether they are free or bound. This necessarily entails some duplication of
effort as well as some transformation machinery between the two representations
that contribute to the complexity of the formalization.
%
Other channel representations are not strong complexity indicators. Note that
the adoption of co-de Bruijn syntax implies the use of context splitting, hence
the two aspects are not completely independent.

There is no strong evidence that the intrinsically typed representation of
processes reduces the size of the formalization. As observed in
\Cref{sec:process-agda}, this choice may help reducing the overall number of
datatypes to be defined, but the definitions are also more involved because they
incorporate invariants and the proofs of typing preservation. In the end, it
seems plausible to speculate that the effort for representing processes, types
and typing rules is not substantially different overall.

Finally, it seems that the use of (explicit) continuations helps taming the
complexity of the formalization more than anything else. Indeed, the six
smallest formalizations (excluding the deadlock freedom results) -- with an
average size of around 62kb -- are all based on continuations, whether they are
explicit (\LinearPI, \SCP, \LCP) or ``virtual'' (\GV), while the remaining ones
are 263kb on average. This observation is in agreement with the considerations
we made earlier in this paper concerning the difference between \emph{updating}
bindings versus \emph{removing and adding} bindings in a typing context. The
adoption of continuations enables a clean linear management of channels and
avoids the need for strong updates on the typing context.

>>>>

We have compared our formalisation with others that have appeared in the
literature and for which we were able to retrieve the source code. We have
looked at those formalisations that include the following features:
\begin{itemize}
\item reduction/evaluation semantics for a calculus with support for binary
    sessions, either natively or through their encoding in terms of linear
    channels~\cite{KobayashiPierceTurner99,Kobayashi02b};
\item a type system along with the usual typing preservation results;
\item optionally, a proof of global progress/deadlock freedom.
\end{itemize}

\citet{GotoEtAl16} describe the formalisation of a session-based variant of the
$\pi$-calculus which supports channel polymorphism. This is the oldest
formalisation of a session-based calculus for which we were able to retrieve the
source code (the work of \citet{Gay01} predates this one, but its source code is
not publicly available any more).

\citet{Thiemann19} formalises a subset of \GV~\citep{GayVasconcelos10}, a
functional language extended with session communication primitives, along with
an interpreter. \citet{CicconePadovani20} have taken inspiration from Thiemann's
work to formalise a variant of the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99} that supports dependent types, so
as to enable the description of communication protocols whose structure depends
on the content of messages.

\citet{CastroFerreiraYoshida20} describe EMTST, a library for the formalisation
of session type systems which includes as case studies the session calculus of
\citet{HondaVasconcelosKubo98} (called ``original system'') and a revised
version of it that is more amenable to be formalised using a locally nameless
representation of channels.

\citet{RouvoetPoulsenKrebbersVisser20} present a library of abstractions
inspired to separation logic aiding the formalisation of interpreters for
computing models supported by linear type systems. One of the presented case
studies is the formalisation of a fragment of \GV~\citep{Wadler14}. The
separating conjunction we have introduced in \Cref{sec:context-agda} and the
typing of the constructors for the representation of processes in
\Cref{sec:process-agda} have been heavily inspired by this work.

% as case study for their approach based on separation logic, which we have
% partially taken inspiration from for the representation of processes
% (\cref{sec:context-agda,sec:process-agda}). Their formalisation adopts the same
% intrinsically-typed representation of processes with nameless channels as our
% own, but their formalisation focuses on the implementation of an interpreter (no
% small-step semantics is defined) and subject reduction is

All the formalisations mentioned so far make use of context splitting. In
contrast, \citet{ZalakainDardha21} formalise a generalisation of the linear
$\pi$-calculus which is parametric in a \emph{usage algebra} (to account for
channel sharing/linearity) and that is based on \emph{leftover
typing}~\cite{Allais17}. 
%
With leftover typing a process is typed with respect to an \emph{input context},
which describes the overall available channels, and an \emph{output context},
which describes the residual channels not consumed by the process. In this way,
it is possible to ``concatenate'' typing judgments instead of splitting typing
contexts beforehand. As \citet{ZalakainDardha21} nicely summarise, ``context
splits contain usage information that is already present'' in processes, and
therefore can be omitted. This is true, but only provided that channels are
given a \emph{name} in processes (\citet{ZalakainDardha21} use de Bruijn indices
to this aim). In fact, the co-de Bruijn representation of
processes~\cite{McBride18}, whereby channels are nameless and context splitting
is performed eagerly, can be seen as the ``dual approach'' of leftover typing:
channel names provide information that is already present in their (singleton)
typing context, hence they can be omitted from contexts and processes.

Another work based on context splitting is the one by
\citet{JacobsBalzerKrebbers22}, who formalise a library of \emph{connectivity
graphs} for reasoning and enforcing deadlock freedom in a variant of
\GV~\citep{GayVasconcelos10,Wadler14}. This is the first formalisation of
deadlock freedom for a calculus of sessions.

@@

    \caption{\label{tab:comparison} A summary of different formalisations of binary session calculi. All sizes are in kb.}

@@

        \textbf{Formalisation} &

@@

\Cref{tab:comparison} summarises the comparison of our formalisation with the
aforementioned ones. The three leftmost colums list the reference papers (sorted
by publication date), the proof assistant and the formalised calculus.
%
For simplicity, we have grouped the calculi into classes identified by an
acronym. Some acronyms, like \CP~\cite{Wadler14} and
\GV~\cite{GayVasconcelos10,Wadler14} are well known in the literature on session
types. It should be noted that the actual calculus being formalised usually
differs from (typically, is a strict subset of) the one identified by the
acronym, but it retains at least some of its key features. We use \SessionPI to
denote (variants of) the session-based $\pi$-calculus presented by
\citet{HondaVasconcelosKubo98} and \LinearPI to denote (variants of) the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99}, into which binary sessions can be
encoded using explicit continuation channels as described by \citet[extended
version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}. Finally, we use
\SCP for the structural version of \CP introduced by
\citet{SanoKavanaghPientka23} and \Calculus for our calculus, a linear version
of \CP with explicit continuations.

The four middle columns of \Cref{tab:comparison} report the relevant qualitative
aspects of the formalisations, namely the technique used to enforce the linear
usage of channels, the technique used to encode channels, whether the
formalisation is intrinsically or extrinsically typed, and whether the calculus
makes use of continuation channels. Concerning the use of continuations, the
approaches based directly on the linear $\pi$-calculus (into which sessions can
be encoded) are marked with {\YES} and those based on a calculus with native
sessions are marked with {\NO}. The calculus \SCP is marked with {\MAYBE}
because, while not being strictly based on the linear $\pi$-calculus, it makes
use of explicit continuations for defining the linearity predicates that check
the linear usage of channels. Finally, all the approaches based on \GV are also
marked with \MAYBE. In principle, \emph{none} of these calculi makes use of
continuations. However, \GV is designed in such a way that each session
operation acting on a channel $s$ returns the result of the operation (if any)
\emph{along with the same channel $s$}. In this way, the type of $s$ can be
``updated'' to take into account the effect of the operation. As observed by
\citet{Padovani17}, this way of modeling the communication primitives is
virtually indistinguishable from one making use of explicit continuation
channels.

The rightmost columns of \Cref{tab:comparison} report the \emph{size} (in
kilobytes) of the formalisations as rough (and possibly questionable) estimates
of their complexity. Instead of counting line numbers, which may be affected by
code indentation styles and syntactical constraints of the proof assistant being
used, we have counted the number of characters after comments have been removed
and spaces have been squeezed.\footnote{Sequences of two or more consecutive
space-like characters are collapsed into a single space. The squeezing is
obtained by running the command \texttt{tr -s [:space:] file} on Unix-like
systems.}
%
The reported sizes account for the source code of the formalisation excluding
examples and any safety result, if present. We have noticed that the
interpretation of ``safety'' varies widely among the formalisations and, except
for our own, they all differ from the one stated in the linearity
challenge~\citep{CarboneEtAl24} when available.
%
Some formalisation define \emph{libraries} which could be reused in different
contexts. In these cases, the size of the library is reported separately next to
the size of the part of the development that uses it.

In general, it is difficult to draw firm conclusions on the effectiveness of the
various approaches in addressing the linearity challenge because the
formalisations differ widely for a variety of entangled factors including the
reference calculus, supported features, proof assistant,
process/expression/channel representation, handling of linearity. Also, most
formalisations predate the publication of the linearity
challenge~\cite{CarboneEtAl24}, hence they are not necessarily targeted to
solving that particular problem. However, some of
them~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,SanoKavanaghPientka23,ZackonEtAl25}
are explicitly motivated by the complexity arising from the linear usage of
resources/channels and from context splitting in particular.

Based on the available data, we can make the following observations.
%
The use of context splitting, which is very well represented, does not seem to
be a good indicator of the complexity of the formalisation, since it spans the
whole range of sizes. Our formalisation is based on context splitting and it is
the only one that covers \CP in full, yet it is by far the smallest one.

The largest formalisations~\cite{GotoEtAl16,CastroFerreiraYoshida20} are also
the ones that adopt a locally nameless representation of channels. In this
formalisations channels are represented in two different ways, depending on
whether they are free or bound. This necessarily entails some duplication of
effort as well as some transformation machinery between the two representations
that contribute to the complexity of the formalisation.
%
Other channel representations are not strong complexity indicators. Note that
the adoption of co-de Bruijn syntax implies the use of context splitting, hence
the two aspects are not completely independent.

There is no strong evidence that the intrinsically typed representation of
processes reduces the size of the formalisation. As observed in
\Cref{sec:process-agda}, this choice may help reducing the overall number of
datatypes to be defined, but the definitions are also more involved because they
incorporate invariants and the proofs of typing preservation. In the end, it
seems plausible to speculate that the effort for representing processes, types
and typing rules is not substantially different overall.

Finally, it seems that the use of (explicit) continuations helps taming the
complexity of the formalisation more than anything else. Indeed, the six
smallest formalisations (excluding the deadlock freedom results) -- with an
average size of around 62kb -- are all based on continuations, whether they are
explicit (\LinearPI, \SCP, \LCP) or ``virtual'' (\GV), while the remaining ones
are 263kb on average. This observation is in agreement with the considerations
we made earlier in this paper concerning the difference between \emph{updating}
bindings versus \emph{removing and adding} bindings in a typing context. The
adoption of continuations enables a clean linear management of channels and
avoids the need for strong updates on the typing context.
